---
title: "Ningaloo ETL"
author: "Florian Mayer"
date: "4 July 2016"
output: html_document
---

# Install
Install required packages by running `setup.R` once in your Ubuntu environment.
```{r, eval=FALSE}
system("sudo apt-get install mdbtools")
install.packages(c("Hmisc", "dplyr", "devtools"))
devtools::install_github("ropensci/ckanr")
```

# Setup
Configure `ckanr` for use with our data catalogue.
The file `setup_ckanr.R` contains the confidential CKAN API key, which gives
the owner's write permissions: `ckanr::ckanr_setup(url=CKAN, key=APIKEY)`.

Create your own `setup_ckanr.R` from the template `setup_ckanr_template.R`.

```{r}
require(Hmisc)
require(dplyr)
require(ckanr)
source("setup_ckanr.R")
```

# Extract data
Donwload and open the Access mdb file from the data catalogue.
```{r, echo=T}
temp <- tempfile()
download.file(resource_show(MDB_RID)$url,temp)
con <- mdb.get(temp, dateformat='%Y-%m-%d')
glimpse(con$tblDBAreaSurveyed)
```

# Transform
TODO: flatten raw data, replicate views

# Load
Upload the output to the data catalogue.
```{r}
# ckanr::resource_update(ETL_RID, path="ningaloo-etl.html")
```

# Cleanup
Close the temporary file.
```{r}
unlink(temp)
```
