---
title: "Turtle Tracks"
author: "Florian Mayer"
date: "19 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(Hmisc)
library(tidyr)
require(dplyr)
require(lubridate)
require(ckanr)
require(DT)
require(mapview)
```

## Using downloaded data

Download CSV from the [Google Fusion Table](https://fusiontables.google.com/DataSource?docid=1wL_dSRNuUCyukJjiUo8RDvFQ0ejWoRpJo2p3S5Rm#map:id=6)
of track counts.

Timestamps of observation date and time are stored in local time by the data collection app. 
We extract the date (dmY) from the timestamp `observation_start_time` into a new 
variable `observation_date`.

Test observations before the start of field work are excluded.

If given, the first of each nest's damage obsevations is shown. If a nest has
several damage observations, the additional observations beyond the first one
will turn up in a separate table, but now shown here.

The basic structure is one record (row) per encountered track (with or without 
nest):

```{r loaddata}
ord <- c("mdYHMOS")
utc <- "UTC"
gmt08 <- "Australia/Perth"

eve_of_fieldwork <- "2016-11-18"

d <- read.csv("data/TrackCount 0.10View.csv", header=T, as.is=T)%>% 
  mutate(
    observation_date=as_date(
      with_tz(
        parse_date_time(observation_start_time, orders=ord, tz=utc)), gmt08)) %>%
  filter(observation_date>eve_of_fieldwork)
glimpse(d)
```

## Plotting nesting effort
The aim of this section is to visualise the sum of observed fresh and old tracks,
grouped by species, and facet-wrapped by sector (pending).

Observations tallied by observation date, species, and nest age:

```{r tally}
tally_data <- d %>%
  group_by(observation_date, species, nest_age) %>%
  tally(sort=F) %>%
  ungroup()
DT::datatable(tally_data)
```


```{r}
ggplot(tally_data, aes(x=observation_date, y=n, colour=species)) +
  geom_point() + facet_wrap(~nest_age)
```


Observations of fresh tracks or nests, 
tallied by observation date, species, and nest type:
```{r}
tally_fresh <- d %>%
  filter(nest_age=="fresh") %>%
  group_by(observation_date, species, nest_type) %>%
  tally(sort=F) %>%
  ungroup()
DT::datatable(tally_fresh)
```

```{r}
ggplot(tally_fresh, aes(x=observation_date, y=n, colour=nest_type)) +
  geom_point() + facet_wrap(~species)
```

**Note** Track and nest counts within the tagging sector (Jetty to Chevron) 
from the night of 20/11/2016 to 21/11/2016 were not recorded
in the ODK Collect app the morning after, as wind and foot traffic during the night
rendered the tracks unreadable. The tracks and nests were recorded on paper data
sheets and will turn up in the tagging database WAMTRAM2.
